Signed Peripherality Functions for Multivariate Statistical Inference

### 2018-03-23 MS-new/4_SDR.tex
main paper v1 done, proofs and intro remain

In the presence of a vector of univariate responses, say $\bfY = (Y_1,Y_2,...,Y_n)^T$, there is substantial literature devoted to utilizing the subspace generated by the basis of $Cov(\bfX)$ in modelling $E(Y|\bfX)$. This ranges from the simple Principal Components Regression (PCR) to Partial Least Squares (PLS) and Envelope methods \citep{Cook10}. Here we concentrate on robust inference using Sufficient Dimension Reduction (SDR) \citep{AdragniCook09}, mainly because it provides a general framework for reducing dimensionality of data directly using top eigenvectors of the covariance matrix of $X$ (albeit in a different manner than PCR) or an appropriate affine transformation of it.

SDR attempts to find out a linear transformation $R$ on $\bfX$ such that $E(Y|\bfX) = E(Y|R(\bfX))$. Assuming that $R(\bfX)$ takes values in $\mathbb R^d, d \leq \min(n,p)$, this can be achieved through an inverse regression model:
%
\begin{equation}
\bfX_y = \bar \bfmu + \Gamma \bfv_y + \bfepsilon
\end{equation}
%
where $\bfX_y = \bfX|Y=y, \bar\bfmu = E\bfX$, $\Gamma$ is a $p \times d$ semi-orthogonal basis for $\mathcal S_\Gamma$, the spanning subspace of $\{ E \bfX_y - \bar\bfmu | y \in S _Y \}$ ($S_y$ is sample space of $Y$) and $\bfv_y = (\Gamma^T \Gamma)^{-1} \Gamma^T (E \bfX_y - \bar\bfmu) \in \mathbb R^d$. The random error term $\bfepsilon$ follows a multivariate normal distribution with mean ${\bf 0}_p$ and covariance matrix $\Delta$. This formulation is straightforward to implement when $Y$ is categorical, while for continuous responses, the vector $\bfy$ is divided into a number of slices.

Under this model the minimal sufficient transformation is $R(\bfX) = \Gamma^T \Delta^{-1} \bfX$. The simplest case of this model is when $\Delta = \sigma^2 I_p$, for which the maximum likelihood estimator of $ R(\bfX)$ turns out to be the first $d$ PCs of $Cov(\bfX)$. Taking $\hat E\bfX_y = \bar\bfX_y$ and $\hat{\bar\bfmu} = \bar\bfX$, one can now estimate $\sigma^2$ as: $\hat\sigma^2 = \sum_{i=1}^p s_{ii}/p$, where $s_{ii}$ is the $i^\text{th}$ diagonal element of $\hat{Cov}_Y( \bfX_Y - \bar\bfX - \hat\Gamma \hat\bfv_Y)$. Following this, predictions for a new observation $\bfx$ is obtained as a weighted sum of the responses:
%
$$
\hat E(Y|\bfX=\bfx) = \frac{\sum_{i=1}^n w_i Y_i}{\sum_{i=1}^n w_i}; \quad w_i = \exp \left[ -\frac{1}{\hat\sigma^2}  \| \hat\Gamma^T (\bfx - \bfX_i) \|^2 \right]
$$
%

We formulate a robust version of the above procedure by estimating the quantities $\Gamma, \bar\bfmu, \bfmu_y, \sigma^2$ by robust methods. Specifically, we take:
%
\begin{itemize}
\item $\tilde \Gamma = $ first $d$ eigenvectors of the sample DCM;
%
\item $\tilde{\bar\bfmu} = $ spatial median of the rows of $X$;
%
\item $\tilde \bfmu_y = $ spatial median of the rows of $(X|Y=y)$, for all $y \in S_Y$;
%
\item $\tilde\sigma^2 = \sum_{i=1}^p [\widehat{\text{MAD}}_Y (X_{Y,i} - \tilde{\bar\mu}_i - \tilde\bfgamma_i^T \tilde\bfv_Y)]^2/p$, with $\tilde\Gamma = (\tilde\bfgamma_1, ..., \tilde\bfgamma_p)^T$.
\end{itemize}
%
The following simulation study using the same setup as in \citep{AdragniCook09} compares the performance of our robust SDR with the original method with or without the presence of bad leverage points in the covariate matrix $X$. For a fixed dimension $p$, we take $n=200, d=1$, generate the responses $Y$ as independent standard normal, and the predictors as $\bfX_Y = \bfgamma^* v_Y^* + \bfepsilon$, with $\bfgamma^*_{p\times 1} = (1,...,1)^T, v_Y = Y + Y^2 + Y^3$ and $Var(\bfepsilon) = 25 I_p$. We measure performance of both SDR models by their mean squared prediction error on another set of 200 observations $(Y^*, \bfX^*)$ generated similarly, and taking the average of these errors on 100 such training-test pair of datasets. Finally we repeat the whole setup for different choices of $p = 5,10,25,50,75,100,125,150$.

@@

Panel (a) of figure \ref{fig:SDRfig} compares prediction errors using robust and maximum likelihood SDR estimates when $X$ contains no outliers, and the two methods are virtually indistinguishable. We now introduce outliers in each of the 100 datasets by adding 100 to first $p/5$ coordinates of the first 10 observations in $X$, and repeat the analysis. Panel (b) of the figure shows that although our robust method performs slightly worse than the case when there were no outliers, it remains more accurate in predicting our of sample observations for all values of $p$.

>>>>

In the presence of a vector of univariate responses, say $\bfy = (y_1,\ldots, y_n)^T$, each $y_i$ being i.i.d. relalizations of a univariate random variable $Y$ that takes values in $\cY \subseteq \BR$, there is substantial literature devoted to utilizing the subspace generated by the basis of $\Sigma = \BV \bfX$ in modelling $\BE(Y|\bfX)$. This ranges from the simple Principal Components Regression (PCR) to Partial Least Squares (PLS) and Envelope methods \citep{Cook10}. In this section, we concentrate on robust inference using Sufficient Dimension Reduction (SDR) \citep{AdragniCook09}, mainly because it provides a general framework for reducing dimensionality of data directly using top eigenvectors of $\Sigma$ (albeit in a different manner than PCR) or an appropriate affine transformation of it.

SDR attempts to find out a linear transformation $\bfR$ on $\bfX$ such that $\BE(Y|\bfX) = \BE(Y| \BR(\bfX))$. Assuming that $\bfR(\bfX)$ takes values in $\mathbb R^d, d \leq \min(n,p)$, this can be achieved through an inverse regression model:
%
\begin{equation}
\bfX_y = \bfmu + \Gamma_d \bfv_y + \bfepsilon,
\end{equation}
%
where $\bfX_y = \bfX|Y=y$, $\Gamma_d \in \BR^{p \times d}$ is a semi-orthogonal basis for $\cS_\Gamma$, the spanning subspace of $\{ \BE \bfX_y - \bfmu | y \in \cY \}$, $\bfv_y = (\Gamma_d^T \Gamma_d)^{-1} \Gamma_d^T (\BE \bfX_y - \bfmu) \in \mathbb R^d$, and $\bfepsilon \sim \cN_p ({\bf 0}, \Delta)$ for some positive-definite matix $\Delta$. This formulation is straightforward to implement when $Y$ is categorical, while for continuous responses, the vector $\bfy$ is divided into a number of slices.

Under this model the minimal sufficient transformation is $\bfR(\bfX) = \Gamma_d^T \Delta^{-1} \bfX$ \citep{AdragniCook09}. The simplest case of this model is when $\Delta = \sigma^2 \BI_p$, for which the maximum likelihood estimator of $ \bfR(\bfX)$ turns out to be the first $d$ PCs of $\Sigma$. Taking $\hat \bfmu_y = \sum_{i=1}^n \bfx_i \BI \{y_i = y\}/n$ and $\hat \bfmu = \sum_{i=1}^n \bfx_i /n$, one can now estimate $\sigma^2$ as: $\hat\sigma^2 = \sum_{i=1}^p s_i /p$, where $s_i$ is the $i^\text{th}$ diagonal element of $\BC_y^T \BC_y/n$, with
%
$$
\BC_y = (\bfc_{y1}, \ldots, \bfc_{yn_y})^T; \quad
n_y = \sum_{i=1}^n \BI \{ y_i = y \},
\bfc_{yi} = \bfx_{yi} - \hat \bfmu - \hat\Gamma_d (\hat \Gamma_d^T \hat \Gamma_d)^{-1} \hat \Gamma_d^T (\hat \bfmu_y - \hat \bfmu ).
$$
%
Following this, predictions for a new observation $\bfx$ is obtained as a weighted sum of the responses:
%
$$
\hat \BE(Y|\bfX=\bfx) = \frac{\sum_{i=1}^n w_i y_i}{\sum_{i=1}^n w_i}; \quad w_i = \exp \left[ -\frac{1}{\hat\sigma^2}  \| \hat\Gamma_d^T (\bfx - \bfx_i) \|^2 \right].
$$
%

We formulate a robust version of the above procedure by estimating the quantities $\Gamma_d, \bfmu, \bfmu_y, \sigma^2$ by robust methods. Specifically, we take:
%
\begin{itemize}
\item $\tilde \Gamma_d = $ first $d$ eigenvectors of the sample DCM;
%
\item $\tilde \bfmu = $ spatial median of the rows of $\BX$;
%
\item $\tilde \bfmu_y = $ spatial median of the rows of $\BX_y$, for each $y \in \cY$;
%
\item $\tilde\sigma^2 = \sum_{i=1}^p \tilde s_i/p$, where $\tilde s_i$ is the $i^\text{th}$ diagonal element of $\tilde \BC_y^T \tilde \BC_y/n$, $\tilde \BC_y$ being the matrix composed of rows $\bfx_{yi} - \tilde \bfmu - \tilde \Gamma_d (\tilde \Gamma_d^T \tilde \Gamma_d)^{-1} \tilde \Gamma_d^T (\tilde \bfmu_y - \tilde \bfmu )$ for $i \in \{ 1, \ldots, n \}$ such that $y_i = y$.
\end{itemize}
%
We compare the performance of our robust SDR with the original method of \cite{AdragniCook09} with or without the presence of bad leverage points in $\Sigma$ using a simulation study. For a given value of $p$, we take $n=200, d=1$, generate the responses $y_1, \ldots, y_n$ as independent standard normal, and the predictors as $\bfX_y = {\bf 1} (y + y^2 + y^3) + \bfepsilon$, with ${\bf 1} \in \BR^p$ consisting of all 1's, and $\BV(\bfepsilon) = 25 \BI_p$. We measure performances of the SDR models by their mean squared prediction error on another set of 200 observations $(\bfy^*, \BX^*)$ generated similarly, and taking the average of these errors on 100 such training-test pairs of datasets. Finally we repeat the full setup for different choices of $p = 5,10,25,50,75,100,125,150$.

@@

Panel (a) of figure \ref{fig:SDRfig} compares prediction errors using robust and maximum likelihood SDR estimates when $\BX$ contains no outliers: here the two methods are virtually indistinguishable. We now introduce outliers in each of the 100 datasets by adding 100 to first $p/5$ coordinates of the first 10 observations in $\BX$, and repeat the analysis. Panel (b) of the figure shows that the robust SDR method remains more accurate in predicting out of sample observations for all values of $p$ than standard SDR.

### 2018-03-23 MS-new/5_Functional.tex
main paper v1 done, proofs and intro remain

\section{Robust inference with functional data}

{\colrbf (Some technical notations)}

We use the approach of \cite{BoenteBarrera15} for performing robust PCA on functional data. Given data on $n$ functions, say $f_1, f_2, ..., f_n \in L^2[0,1]$, each observed at a set of common design points $\{ t_1, ..., t_m \} $, we model each function as a linear combination of $p$ mutually orthogonal B-spline basis functions $\delta_1, ..., \delta_p$. Following this, we map data for each of the functions onto the coordinate system formed by the spline basis:
%
\begin{equation}
\tilde x_{ij} = \sum_{l=2}^m f_i(t_l) \delta_j(t_l) (t_l - t_{l-1}); \quad 1 \leq i \leq n, 1 \leq j \leq p
\end{equation}
%
We now do depth-based PCA on the transformed $n \times p$ data matrix $\tilde X$, and obtain the rank-$q$ approximation ($q \leq p$) of the $i^\text{th}$ observation using the robust $p \times q$ loading matrix $\tilde P$ and robust $q \times 1$ score vector $\tilde\bfs_i$:
%
$$ \widehat{\tilde\bfx_i} = \tilde\bfmu + \tilde P \tilde \bfs_i $$
%
with $\tilde\bfmu$ being the spatial median of $\tilde X$. Then we transform this approximation back to the original coordinates: $\hat f_i (t_l) = \sum_{j=1}^p \widehat{ \tilde x}_{ij} \delta_j (t_l)$.

Detection of anomalous observations is of importance in real-life problems involving functional data analysis. We now demonstrate the utility of our robust method for detecting functional outliers through two data examples.

\textbf{(SD and OD definition, cutoffs... from previous manuscript)
}

\begin{figure}[t]
%\captionsetup{justification=centering, font=footnotesize}
\begin{center}
\subfigure[]{\epsfxsize=.32\linewidth\epsfbox{../Codes/Elnino_functional1}}
\subfigure[]{\epsfxsize=.32\linewidth\epsfbox{../Codes/Elnino_functional2}}
\subfigure[]{\epsfxsize=.32\linewidth\epsfbox{../Codes/Elnino_functional3}}\\
\vspace{-1em}
\subfigure[]{\epsfxsize=.32\linewidth\epsfbox{../Codes/Octane_functional1}}
\subfigure[]{\epsfxsize=.32\linewidth\epsfbox{../Codes/Octane_functional2}}
\subfigure[]{\epsfxsize=.32\linewidth\epsfbox{../Codes/Octane_functional3}}
%\subfigure[]{\epsfxsize=0.35\linewidth \epsfbox{../Codes/SDRcomparison_out}}
\caption{Actual sample curves, their spline approximations and diagnostic plots respectively for El-Ni\~no (a-c) and Octane (d-f) datasets}
\label{fig:fPCAfig}
\end{center}
\end{figure}

We first look into the El-Ni\~no data, which is part of a larger dataset on potential factors behind El-Ni\~no oscillations in the tropical pacific available in \url{http://www.cpc.ncep.noaa.gov/data/indices/}. This records monthly average Sea Surface Temperatures from June 1970 to May 2004, and the yearly oscillations follow more or less the same pattern (see panel a of figure \ref{fig:fPCAfig}). Using a cubic spline basis with knots at alternate months starting in June gives a close approximation of the yearly time series data (panel b), and performing depth-based PCA with $q=1$ results in two points having their SD and OD larger than cutoff (panel c). These points correspond to the time periods June 1982 to May 1983 and June 1997 to May 1998 are marked by black curves in panels a and b), and pinpoint the two seasons with strongest El-Ni\~no events.

Our second application is on the Octane data, which consists of 226 variables and 39 observations \citep{esbensen94}. Each sample is a gasoline compound with a certain octane number, and has its NIR absorbance spectra measured in 2 nm intervals between 1100 - 1550 nm. There are 6 outliers here: compounds 25, 26 and 36-39, which contain alcohol. We use the same basis structure as the one in El-Ni\~no data here, and again the top robust PC turns out to be sufficient in identifying all 6 outliers (panels d, e and f of figure \ref{fig:fPCAfig}).

>>>>

\section{Robust inference with functional data}
\label{section:fpcaSection}
Detection of anomalous observations is of importance in real-life problems involving functional data analysis, and functional PCA is a widely used tool in this setting. In this section we utilize robust principal components from the DCM for this purpose. We use the approach of \cite{BoenteBarrera15} for performing robust PCA on functional data using the estimated PCs from the DCM. Here we have a data matrix $\BH$, that stores the values of a set of $n$ curves, say $ \cF = \{ f_1, \ldots , f_n  \} \in L^2[0,1]$, each observed at a set of common design points $\{ t_1, ..., t_m \} $. We model each of these functions as a linear combination of $p$ mutually orthogonal B-spline basis functions $\cD = \{ \delta_1, ..., \delta_p \}$. Following this, we map data for each of the functions onto the coordinate system formed by the spline basis:
%
\begin{equation}
T( \BH; \cF, \cD)_{ij} = \sum_{l=2}^m f_i(t_l) \delta_j(t_l) (t_l - t_{l-1}); \quad 1 \leq i \leq n, 1 \leq j \leq p.
\end{equation}
%
We now do depth-based PCA on the transformed $n \times p$ data matrix $T(\BH; \cF, \cD) \equiv T(\BH)$, and obtain the rank-$q$ approximation ($q \leq p$) of the $i\Th$ observation using the robust $p \times q$ loading matrix $\tilde \bfP$ and robust $q \times 1$ score vector $\tilde\bfs_i$:
%
$$
\widehat T(\BH)_i = \tilde \bfmu + \tilde \bfP \tilde \bfs_i,
$$
%
with $\tilde \bfmu$ being the spatial median of $T(\BH)$. Then we transform this approximation back to the original coordinates: $\hat f_i (t_l) = \sum_{j=1}^p \widehat T(\BH)_{ij} \delta_j (t_l)$.

We demonstrate the utility of our robust method for detecting functional outliers through two data examples. For any method of PCA with $k$ components on a dataset of $n$ observations and $p$ variables, the \textit{score distance} (SD) and \textit{orthogonal distance} (OD) for $i^\text{th}$ observation ($i=1,2,...,n$) are defined as:
%
$$
SD_i = \sqrt{ \sum_{j=1}^k \frac{s^2_{ij}}{\hat \lambda_j}}; \quad OD_i = \| \bfx_i - \bfP\bfs_i^T \|,
$$
%
where $\bfs_1, \ldots ,\bfs_n$ are the score vectors, $\bfP \in \BR^{p\times k}$ the loading matrix, and $\hat \lambda_1,\ldots ,\hat \lambda_k$ are eigenvalues obtained from the PCA. From a practical standpoint, $SD_i$ can be interpreted as a weighted norm of the projection of the $i^\text{th}$ point on the hyperplane formed by first $k$ principal components, and $OD_i$ the orthogonal distance of point $i$ from that hyperplane. For outlier detection, following \cite{hubert05} we set the upper cutoff values for score distances at $(\chi^2_{2,.975})^{1/2}$ and orthogonal distances at $[\text{median}(OD^{2/3}) + \text{MAD}(OD^{2/3})\Phi^{-1}(0.975)]^{3/2}$, where $\Phi(.)$ is the standard normal cumulative distribution function.

\begin{figure}
%\captionsetup{justification=centering, font=footnotesize}
\begin{center}
\subfigure[]{\epsfxsize=.35\linewidth\epsfbox{../Codes/Elnino_functional1}}
\subfigure[]{\epsfxsize=.35\linewidth\epsfbox{../Codes/Octane_functional1}}\\

\subfigure[]{\epsfxsize=.35\linewidth\epsfbox{../Codes/Elnino_functional2}}
\subfigure[]{\epsfxsize=.35\linewidth\epsfbox{../Codes/Octane_functional2}}\\

\subfigure[]{\epsfxsize=.35\linewidth\epsfbox{../Codes/Elnino_functional3}}
\subfigure[]{\epsfxsize=.35\linewidth\epsfbox{../Codes/Octane_functional3}}\\
%\subfigure[]{\epsfxsize=0.35\linewidth \epsfbox{../Codes/SDRcomparison_out}}
\caption{Actual sample curves, their spline approximations and diagnostic plots respectively for El-Ni\~no (a,c,e) and Octane (b,d,f) datasets}
\label{fig:fPCAfig}
\end{center}
\end{figure}

We consider the El-Ni\~no data, which is part of a larger dataset on potential factors behind El-Ni\~no oscillations in the tropical pacific available in \url{http://www.cpc.ncep.noaa.gov/data/indices}, as the first test case for outlier detection using our robust functional PCA. This records monthly average Sea Surface Temperatures from June 1970 to May 2004, and the yearly oscillations follow more or less the same pattern (see panel a of Figure~\ref{fig:fPCAfig}). Using a cubic spline basis with knots at alternate months starting in June gives a close approximation of the yearly time series data (panel c), and performing depth-based PCA with $q=1$ results in two points having their SD and OD larger than cutoff (panel e). These points correspond to the time periods June 1982 to May 1983 and June 1997 to May 1998 are marked by black curves in panels a and c, and pinpoint the two seasons with strongest El-Ni\~no events.

Our second application is on the Octane data, which consists of 226 variables and 39 observations \citep{esbensen94}. Each sample is a gasoline compound with a certain octane number, and has its NIR absorbance spectra measured in 2 nm intervals between 1100 - 1550 nm. There are 6 outliers here: compounds 25, 26 and 36-39, which contain alcohol. We use the same basis structure as the one in El-Ni\~no data here, and again the top robust PC turns out to be sufficient in identifying all 6 outliers (panels b, d and f of Figure~\ref{fig:fPCAfig}).

### 2018-03-23 MS-new/6_Conclusion.tex
main paper v1 done, proofs and intro remain

\section{Conclusion}\label{section:sec7}

In the above sections we introduce a covariance matrix based on depth-based multivariate ranks that keeps the eigenvectors of the actual population unchanged for elliptical distributions. We provide asymptotic results for the sample DCM, its eigenvalues and eigenvectors. Bounded influence functions as well as simulation studies suggest that the eigenvector estimates obtained from the DCM are highly robust, yet do not lose much in terms of efficiency. Thus it provides a plausible alternative to existing approaches of robust PCA that are based on estimation of covariance matrices (for example SCM, Tyler's scatter matrix, D\"{u}mbgen's symmetrized shape matrix).

{\colrbf (some text)}

>>>>

%\section{Conclusion}\label{section:sec7}
%
%In the above sections we introduce a covariance matrix based on depth-based multivariate ranks that keeps the eigenvectors of the actual population unchanged for elliptical distributions. We provide asymptotic results for the sample DCM, its eigenvalues and eigenvectors. Bounded influence functions as well as simulation studies suggest that the eigenvector estimates obtained from the DCM are highly robust, yet do not lose much in terms of efficiency. Thus it provides a plausible alternative to existing approaches of robust PCA that are based on estimation of covariance matrices (for example SCM, Tyler's scatter matrix, D\"{u}mbgen's symmetrized shape matrix).

\section{Conclusion}\label{section:sec7}
\label{section:Conclusion}
In the above sections we elaborate on a proposed transformation based on the idea of combining sign functions in an inner product space and certain transformations of general peripherality functions defined using probability measures on the same space. Based on the conditions we impose in course of the chapter, we essentially end up using data depths as scalar multiples of the spatial sign in estimation and testing problems for the location parameter of an elliptical distribution in $\BR^p$, and using inverse depth functions for robust estimation of different components of its covariance matrix. As demonstrated by several simulation studies and data examples, in all these situations the use of this composite transformation vector brings about efficiency gains, as well as favorable robustness properties in terms of bounded influence functions and deviations from Gaussianity.

Regarding the multivariate rank transformation we propose, to be noted is the fact that the mapping $\bfX \rightarrow \tilde \bfX$ is in fact one-to-one for elliptical distributions. Thus such rank vectors can possibly be used for inference based on transformation-retransformation type techniques (e.g. \cite{ChakrabortyChaudhuri96,ChakrabortyChaudhuriOja98}). We defer this to future research. Finally, while these rank vectors have excellent intuitive appeal in preserving the shape of a multivariate data cloud, it would be interesting to study the properties of transformations similar to \eqref{eqn:rank} in general Hilbert spaces, as well as explore their applications in different data-analytic domains.
